How to select predictive models for decision-making or causal inference?

### 2024-10-20 response_to_reviewer.md

> The reviewer is right that Eq. (2) is part of constructing the classic
> g-computation estimator, as well as Eq. (1) and Eq. (3). We added a
> sentence at the begining of the section to stress that it is a section
> summarizing a classic framework.

> Eq. (3) is an expression of the estimator of the ATE based on a model of the outcome, sometimes called g-estimation. It is unbiased under consistency (Assumption. 3) and unconfoundedness (Assumption 1). To prove that, take the expectation of Eq. (3).  Then apply  Eq. (1) which is true under consistency (Assumption 3) and unconfoundedness (Assumption 1). Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq. (2).

> G-estimation itself does not control error on both populations. The formulas in Eq. (1-3) are all partly oracle formulas: they rely on conditional expectations but give no specific procedures on how to compute them. We added a clarification on this topic above Eq. (4) and above Eq. (8).

*Furthermore, the Robinson decomposition (R-decomposition) should be better discussed and its role in the context of causal inference should be better commented.
R-decomposition introduces two quantities, the conditional mean outcome and the probability to be treated (known as propensity score). What is the role of the propensity score? How is it calculated? Is it calculated simply by a frequentist approach (number of treated/total number of patients)?*

> G-computation relies on one choice of decomposition of the CATE estimation (Eq. (2)), which leads to into a sequence of regressions, where the only statistical estimate is the response function µa(x)  [Chernozhukov et al., 2024, chapter 15](https://causalml-book.org/). Other decomposition of the outcome exist, as a decomposition is just a mathematical rewriting... The R-decomposition is one way of identifying the causal quantity of interest which introduce two statistical quantity that have to be modelled: the mean outcome from the treatment effect and the propensity score. Any statistical estimator can be chosen to model these quantities. The frequentist approach is one (naive) such estimator. In practice, researchers often estimate the propensity score with a logistic regression, but other methods such as Boosted trees, Random Forests, or Neural Networks can be used. A good ps estimator should be well calibrated but this question is out of scope for this paper. In our discussion, we briefly mentioned this interesting research question on what properties should have a good estimator for the propensity score.

### 2024-10-20 how_to_select_latex.tex
iter

Should we know the two potential outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two potential outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following
equality, linking the target potential quantity to the observed data:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$. Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
However, the g-computation framework presented above is written in terms
of ``perfect'' conditional expectations (oracles), it does not control an
error, \emph{eg} on both populations as highlighted in Figure
\ref{fig:toy_example}. Selection procedures are needed to favor the best model for each dataset.

A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ with observed data $O$ and a splitting strategy of the data for estimating the different regressions (nuisances) involved in the risk.
%
Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}

>>>>

Should we know the two potential outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two potential outcomes can be estimated from observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the following
equality, linking the target quantity to the observed data:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}):
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$. Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
%
However, the g-computation framework presented above is written in terms
of ``perfect'' conditional expectations (oracles), it does not control an
error, \emph{eg} on both populations as highlighted in Figure
\ref{fig:toy_example}. Selection procedures are needed to find the best
conditional-expectation models.

A selection procedure combines a risk $\ell$, evaluating the quality of a
model $f$ with observed data $O$, and a splitting strategy of the data to
estimate different regressions (nuisances) involved in the risk.
%
Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
a family of such estimators. Our goal is to select the best candidate in this
family for the observed dataset $O$ using a risk
$\ell$:
\begin{equation}
    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
    \label{eq:causal_model_selection}
\end{equation}
